Data Migration under a Changing Schema in Ampersand

### 2023-11-25 2022Migration/articleMigrationFoIKS.tex
wip

\item\label{step3}
   The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed,
   and a blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
% \begin{verbatim}
%    RELATION fixed_TOTr[A*A]
%    RULE Block_TOTr : fixed_TOTr |- new.r;new.r~
%    MESSAGE "Relation r[A*B] must remain total."
%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")
% \end{verbatim}
   \begin{align}
      \rels_2={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
      \rules_\text{block}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf with}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}\cap{\tt fixed}_u
      \end{array}\label{eqn:blockRule}\\
      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
   \end{align}
   % Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
\item\label{step4}
   We use a transactional invariant to produce the population of the helper relation ${\tt fixed}_u$.
% \begin{verbatim}
%    ENFORCE fixed_TOTr >: I /\ new.r;new.r~
% \end{verbatim}
   \begin{equation}
     \transactions_2\ =\ \{{\tt fixed}_u\mapsfrom\lambda\dataset.~\cmpl{\viol{\overrightarrow{u}}{\dataset}}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
   \end{equation}
\item\label{step5} To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
% \begin{verbatim}
%    ROLE User MAINTAINS TOTr
%    RULE TOTr : I |- new.r;new.r~
%    MESSAGE "Please, make relation r[A*B] total."
%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
% \end{verbatim}
   \begin{align}
      \busConstraints_\text{fix}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf with}\label{eqn:Bfix}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{\overrightarrow{u}}{\dataset}-{\tt fixed}_u
      \end{array}\\
      &\mid u\in\rules_{\schema'}-\rules_{\schema}\}\notag
   \end{align}
   In some cases, a migration engineer can invent ways to satisfy these invariants automatically.
   For this purpose, the generator must produce source code (as opposed to compiled code) to allow the migration engineer
   to replace a business constraint with transactional invariants of her own making.
   After all violations are fixed, i.e. when equation~\ref{eqn:readyForMoC} is satisfied,
   the migration engineer can switch the ingress to the desired system.
   This occurs at MoC and
   replaces $\rules_\text{block}$ in the migration system by the blocking invariants of the desired system.
   This moment arrives when:
   \begin{align}
      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}\subseteq{\tt fixed}_u
   \label{eqn:readyForMoC}
   \end{align}
   After this, the migration engineer can remove the migration system and the old system.

\item Let us combine the above into a single migration schema:
   \begin{align}
      \schema_\migrsys=\langle{}&\concepts_\dataset\cup\concepts_{\dataset'},\label{eqn:schema migrsys}\\
      &\overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,\notag\\
      &\rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}},\notag\\
      &\transactions_1\cup\transactions_2\cup\overrightarrow{\transactions_{\schema'}},\notag\\
      &\busConstraints_\text{fix}\cup\busConstraints_{\schema'}\rangle\notag
   \end{align}
   This schema represents the migration system.
   In our reasoning, we have only used information from the schemas of the existing system and the desired system.
   This shows that it can be generated from these schemas without using any knowledge of the data sets.
\end{enumerate}

@@

\subsection{Validation}
   The proof of concept gives but one example of something that works.
   To validate that this works in all cases, we must verify that:
\begin{enumerate}
\item The predeployment yields a migration system that copies the designated triples to the desired system in finite time.
\item The migration system satisfies the definition of information system (\ref{def:information system}) at MoT, especially that it has no blocking violations (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
\item The violations that business actors must fix are finite in number and decrease monotonically, so the business can fulfil the condition for the MoC (requirement~\ref{eqn:readyForMoC}) in finite time.
\item The behavior of the migration system and the desired system is identical, so moving the ingress from the migration system to the desired system is not noticable for end-users.
\end{enumerate}
   Let us discuss these points one by one.

   The relations to be copied from the old system are those relations that the desired system retains ($\rels_{\schema'}\cap\rels_{\schema}$).
   For each $r$, $\rels_1$ contains a relation ${\tt copy}_r$ (eqn.~\ref{eqn:copy relations}).
   After the MoT, no more changes will occur in the existing system.
   In other words, the population of every relation in $\rels_{\schema}$ becomes stable and so does every ${\tt copy}_r$.
   When the migration system has copied those relations,
   every ${\tt copy}_r$ is populated and eqn.~\ref{eqn:copyingTerminates}) is true.
   Hence, the migration system will make no more changes to ${\tt copy}_r$ and the transactional invariants in $\transactions_1$ will cause no more changes. Effectively, $\transactions_1$ becomes redundant once the copying is done.

   Then, the definition of the migration system, def.~\ref{dfn:migration system}, contains only blocking invariants that exist in the existing system as well.
   For this reason, all blocking invariants of the migration system are satisfied on MoT.
   Since it contains no other blocking invariants, the migration system satisfies requirement~\ref{eqn:satisfaction}.

   Thirdly, the problematic constraints are the blocking invariants of the desired system that were not in the existing system ($\rules_{\schema'}-\rules_{\schema}$ in def.~\ref{eqn:Bfix}).
   The migration engineer might introduce transactional invariants to satisfy these invariants automatically,
   but otherwise, definition~\ref{eqn:Bfix} defines them as business constraints to be resolved by the business.
   Each violation in those rules that a business actor resolves will never reappear,
   because it is registered in ${\tt fixed}_r$.
   So when all violations are resolved, the constraints in ... have effectively become blocking constraints.
\section{Conclusions}
\label{sct:Conclusions}
   In this paper, we describe the data migration as going from an existing system to a desired one, where the schema changes.
   As Ampersand generates information systems, creating a new system can be a small task, allowing for incremental deployment of new features.
   We describe the parts of a system that have an effect on data pollution.
   We assume that the existing system does not violate any constraints of its schema, but address other forms of data pollution:
   constraints that are not in the schema but are in the desired schema are initially relaxed such that the business can start using the migration system, after which this form of data pollution needs to be addressed by human intervention.
   We propose a method for doing migration such that only a finite amount of human intervention is needed.
   Our method allows a system similar to the desired system to be used while the intervention takes place.

>>>>

\item\label{step3}
   The new blocking invariants are $\rules_{\schema'}-\rules_{\schema}$.
   For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, to register all violations that are fixed,
   and a blocking invariant $v$ in the migration system that blocks fixed violations from recurring:
% \begin{verbatim}
%    RELATION fixed_TOTr[A*A]
%    RULE Block_TOTr : fixed_TOTr |- new.r;new.r~
%    MESSAGE "Relation r[A*B] must remain total."
%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")
% \end{verbatim}
   \begin{align}
      \rels_2={}&\{{\tt fixed}_u\mid u \in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\\
      \rules_\text{block}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf with}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{u}{\dataset}\cap{\tt fixed}_u
      \end{array}\label{eqn:blockRule}\\
      &\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\notag
   \end{align}
   % Note that the blocking invariants from $\schema'\cap\schema$ are also used in $\migrsys$ to ensure continuity.
\item\label{step4}
   We use a transactional invariant to produce the population of the helper relation ${\tt fixed}_u$.
% \begin{verbatim}
%    ENFORCE fixed_TOTr >: I /\ new.r;new.r~
% \end{verbatim}
   \begin{equation}
     \transactions_2\ =\ \{{\tt fixed}_u\mapsfrom\lambda\dataset.~\cmpl{\viol{u}{\dataset}}\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\label{eqn:enforceForRules}
   \end{equation}
\item\label{step5} To signal users that there are violations that need to be fixed, we generate a business constraint for each new blocking invariant $u$:
% \begin{verbatim}
%    ROLE User MAINTAINS TOTr
%    RULE TOTr : I |- new.r;new.r~
%    MESSAGE "Please, make relation r[A*B] total."
%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
% \end{verbatim}
   \begin{align}
      \busConstraints_\text{fix}={}&\{v\ 
      \begin{array}[t]{l}
         \text{\bf with}\label{eqn:Bfix}\\
         \sign{v}=\sign{u}\\
         \viol{v}{\dataset}=\viol{u}{\dataset}-{\tt fixed}_u
      \end{array}\\
      &\mid u\in\overrightarrow{\rules_{\schema'}-\rules_{\schema}}\}\notag
   \end{align}
   In some cases, a migration engineer can invent ways to satisfy these invariants automatically.
   For this purpose, the generator must produce source code (as opposed to compiled code) to allow the migration engineer
   to replace a business constraint with transactional invariants of her own making.
   After all violations are fixed, i.e. when equation~\ref{eqn:readyForMoC} is satisfied,
   the migration engineer can switch the ingress to the desired system.
   This occurs at MoC and
   replaces $\rules_\text{block}$ in the migration system by the blocking invariants of the desired system.
   This moment arrives when:
   \begin{align}
      \forall u\in\rules_{\schema'}-\rules_{\schema}.~\viol{\overrightarrow{u}}{\dataset}\subseteq{\tt fixed}_u
   \label{eqn:readyForMoC}
   \end{align}
   After this, the migration engineer can remove the migration system and the old system.

\item Let us combine the above into a single migration schema:
   \begin{align}
      \schema_\migrsys=\langle{}&\concepts_\dataset\cup\concepts_{\dataset'},\label{eqn:schema migrsys}\\
      &\overleftarrow{\rels_{\schema}}\cup\overrightarrow{\rels_{\schema'}}\cup\rels_1\cup\rels_2,\notag\\
      &\rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}},\notag\\
      &\transactions_1\cup\transactions_2\cup\overrightarrow{\transactions_{\schema'}},\notag\\
      &\busConstraints_\text{fix}\cup\overrightarrow{\busConstraints_{\schema'}}\rangle\notag
   \end{align}
   This schema represents the migration system.
   In our reasoning, we have only used information from the schemas of the existing system and the desired system.
   This shows that it can be generated from these schemas without using any knowledge of the data sets.
\end{enumerate}

@@

\subsection{Validation}
   The proof of concept gives but one example of something that works.
   To validate that this works in all cases, we must verify that:
\begin{enumerate}
\item The predeployment yields a migration system, $\migrsys$, that copies the designated triples to the desired system in finite time.
\item $\migrsys$ satisfies the definition of information system (\ref{def:information system}) at MoT, especially that it has no violations of blocking invariants (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
\item The violations that business actors must fix are finite in number and decrease monotonically, so the business can fulfil the condition for the MoC (requirement~\ref{eqn:readyForMoC}) in finite time.
\item The behavior of $\migrsys$ and the desired system is identical, so moving the ingress from $\migrsys$ to the desired system is not noticable for end-users.
\end{enumerate}
   Let us discuss these points one by one.

   The relations to be copied from the old system are those relations that the desired system retains ($\rels_{\schema'}\cap\rels_{\schema}$).
   For each $r$ to be copied, $\migrsys$ contains a relation ${\tt copy}_r$ in $\rels_1$ (eqn.~\ref{eqn:copy relations}).
   After the MoT, the ingress sends all change events to $\migrsys$, so the existing system can finish the work it is doing for transactional invariants and will not change after that.
   In other words, the population of every relation in $\rels_{\schema}$ becomes stable and so does every ${\tt copy}_r$.
   At that point in time, eqn.~\ref{eqn:copyingTerminates} is true and stays true.
   Effectively, $\transactions_1$ becomes redundant once the copying is done.

   Then, $\migrsys$ contains only blocking invariants that exist in the existing system as well (def.~\ref{dfn:migration system}).
   For this reason, all of these blocking invariants are satisfied on MoT.
   Since $\migrsys$ contains no other blocking invariants, it satisfies requirement~\ref{eqn:satisfaction}.

   Thirdly, the problematic constraints are the blocking invariants of the desired system that were not in the existing system ($\rules_{\schema'}-\rules_{\schema}$ in def.~\ref{eqn:Bfix}).
   The migration engineer might introduce transactional invariants to satisfy these invariants automatically,
   but otherwise, definition~\ref{eqn:Bfix} defines them as business constraints, $\busConstraints_\text{fix}$, to be resolved by the business.
   Each violation in $\busConstraints_\text{fix}$ that a business actor resolves,
   will never reappear because it is registered in ${\tt fixed}_r$ by the transactional invariants of $\transactions_2$ (eqn.~\ref{eqn:enforceForRules}).

   So finally, when all violations are resolved, the constraints in $\busConstraints_\text{fix}$ have effectively become blocking invariants.
   The blocking invariants in the desired system consist of $\rules_{\schema}$ and $\busConstraints_\text{fix}$, which is equivalent to $\rules_{\schema'}$.
   Hence we can replace $\busConstraints_\text{fix}\cup\rules_{\schema}$ with $\rules_{\schema'}$ after condition~\ref{eqn:readyForMoC} is satisfied.

   Now we can assemble the results.
   Every rule in $\rules_\text{block}$ has become blocking, so after the MoC $\rules_\text{block}\cup\overrightarrow{\rules_{\schema}\cap\rules_{\schema'}}=\overrightarrow{\rules_{\schema'}}$ in the migration system, which is $\rules_{\schema'}$ in the desired system.
   Since $\transactions_1$ and $\transactions_2$ are redundant after the MoC,
   we can retain $\overrightarrow{\transactions_{\schema'}}$ in $\migrsys$,
   which is equivalent to $\transactions_{\schema'}$ in the desired system.
   Likewise, $\busConstraints_\text{fix}$ has become redundant, so $\migrsys$ can do with just $\overrightarrow{\busConstraints_{\schema'}}$.
   In the desired system, that is equivalent to $\busConstraints_{\schema'}$.
   So, the constraints in the desired system after the MoC are equivalent to the constraints in the MoC

\section{Conclusions}
\label{sct:Conclusions}
   In this paper, we describe the data migration as going from an existing system to a desired one, where the schema changes.
   As Ampersand generates information systems, creating a new system can be a small task, allowing for incremental deployment of new features.
   We describe the parts of a system that have an effect on data pollution.
   We assume that the existing system does not violate any constraints of its schema, but address other forms of data pollution:
   constraints that are not in the schema but are in the desired schema are initially relaxed such that the business can start using the migration system, after which this form of data pollution needs to be addressed by human intervention.
   We propose a method for doing migration such that only a finite amount of human intervention is needed.
   Our method allows a system similar to the desired system to be used while the intervention takes place.
